Smoothing seismicity techniques applied to seismic source characterization and probabilistic seismic hazard analysis in Brazil

### 2015-05-13 d03-acronyms.tex
fixing typos and cleaning for now
adding completeness biblio, text and figures
completeness and introduction almost finished... figure and bibliography update
start of introduction review

\newacronym{psha}{PSHA}{Probabilistic Seismic Hazard Analysis}
\newacronym{dsha}{DSHA}{Deterministic Seismic Hazard Analysis}
\newacronym{GMPE}{\gls{gmpe}}{\glsdesc{gmpe}}
\newacronym{mfd}{MFD}{Magnitude Frequency Distribution}
\newacronym{gr}{GR}{Gutenberg-Richter}
\newacronym{cnn}{CNN}{Coupled-Nearest-Neighbour}
\newacronym{va}{r.v.}{random variable}
\newacronym{iid}{i.i.d.}{independent and identicaly distributed}
\newacronym{pdf}{density}{probability density function}
\newacronym{pmf}{pmf}{probability mass function}
\newacronym{kde}{KDE}{Kernel Density Estimation}

@@

\newacronym{obsis}{ObSis}{Seismological Observatory}
\newacronym{unb}{UnB}{Brasilia University}
\newacronym{iag}{IAG}{Institute of Astronomy, Geophysics and Atmosferical Sciences}
\newacronym{usp}{USP}{São Paulo University}
\newacronym{ipt}{IPT}{Institute of Technological Research}
\newacronym{unesp}{UNESP}{São Paulo State University}
\newacronym{ufrn}{UFRN}{Federal University of Rio Grande do Norte}

@@

>>>>

\newacronym{psha}{PSHA}{Probabilistic Seismic Hazard Analysis}
\newacronym{dsha}{DSHA}{Deterministic Seismic Hazard Analysis}
\newacronym{GMPE}{\gls{gmpe}}{\glsdesc{gmpe}}
\newacronym{mfd}{MFD}{Magnitude Frequency Distribution}
\newacronym{gr}{GR}{Gutenberg-Richter}
\newacronym{cnn}{CNN}{Coupled-Nearest-Neighbor}
\newacronym{va}{r.v.}{random variable}
\newacronym{iid}{i.i.d.}{independent and identically distributed}
\newacronym{pdf}{density}{probability density function}
\newacronym{pmf}{pmf}{probability mass function}
\newacronym{kde}{KDE}{Kernel Density Estimation}
\newacronym{pde}{PDE}{Probability of Detecting an Earthquake}
\newacronym{tcef}{TCEF}{temporal course of earthquake frequency}

@@

\newacronym{obsis}{ObSis}{Seismological Observatory}
\newacronym{unb}{UnB}{Brasilia University}
\newacronym{iag}{IAG}{Institute of Astronomy, Geophysics and Atmospherical Sciences}
\newacronym{usp}{USP}{São Paulo University}
\newacronym{ipt}{IPT}{Institute of Technological Research}
\newacronym{unesp}{UNESP}{São Paulo State University}
\newacronym{ufrn}{UFRN}{Federal University of Rio Grande do Norte}

@@

\newacronym{relm}{RELM}{Regional Earthquake Likelihood Models}
\newacronym{ed}{ED}{error diagrams}

### 2015-05-13 manuscript.tex
start of introduction review

\begin{abstract}
Intraplate seismicity is a hard problem to solve in tectonics. Brazil, as others intraplate environments, mainly during the lack of observed seismicity the seismic hazard assessment are also difficult and the methods applied are often close correlated to the geometries defined by the experts with valuable criteria.
% Intraplate seismicity is a hard problem to solve in tectonics. In general, the earthquake occurrence is lower than others ``active'' regions. In the Brazilian case, there is a lack of observed seismicity due the short (in geological scale) observation timeframe. In addition, the populational density decreases towards to interior of the country and changes the reporting capabilities historicaly.

% This diffused intraplate seismicity pattern is often modeled as polygons representing different seismic zones or area sources for \gls{psha} purposes. Their geometries are defined by expert opinions.

% Other possible approach to deal this is develop zoneless or smoothed  models for the seismicity, conside
Three smoothed seismicity techniques are reviewed and their application on construct models of Brazilian seismicity and its applications on long-term seismic hazard assessment in Brazil are also discussed. All these smoothing methods are based on a \gls{kde}. The main difference between them is the bandwidth selection process.

@@

All the hazard computations was performed on \gls{oq} \citep{pagani_2014} open source software from \acrfull{gem} Foundation. Until the present moment, no focal mechanism information has been included or considered on these smoothed seismicity models. All hazard computations has been done using a single \gls{GMPE}. A more detailed study on \glspl{GMPE} selection should be developed further and is out of this paper scope.

The results shown that all methods clearly most of the well known seismogenic regions and could be used in the Brazilian seismicity modeling. All methods improve the spatial resolution of the GSHAP hazard model, the only previously available for the region. These results should be considered on future reviews of Brazilian design building codes and procedures.
%Tectonic connections between these seismicity models and stress release could be developed in the future.

@@

\subsection{Intraplate seismicity}

Brazil occupies a large stable continental region with seismicity rate lower than Central and Eastern North America, Australia, India or Northern Europe \citep{giardini_1999,johnston_1996a}. Spanning most of the central region of the South American plate, Brazil has earthquakes above magnitude 6 mb with a return period of 50 years or so. Historical records are very short with almost no data before the XX century. Historical evidence of magnitude a 7 has been reported in the Amazon \citep{berrocal_1984,veloso_2014}. However, as in other intraplate settings, seismic hazard is not negligible due to the low seismic attenuation in the cold and thick lithosphere (magnitude 5 mb can be felt up to ~500 km away \citep{assumpcao_et_al_2014}) and special seismic hazard studies have been required for critical facilities such as nuclear plants \citep{almeida_etal_2013}.

Very few studies related to seismic hazard, such as \citet{berrocal_1996} for SE Brazil, have been carried out so far and the first Brazilian seismic building code \citep{nbr_15421_2006} was based on the GSHAP \citep{giardini_1999} results for South America \citep{tanner_shepherd_1997, shedlock_tanner_1999}. Brazilian seismicity is concentrated in several seismic zones \citep{berrocal_1984, assumpcao_et_al_2014}. Except for a peak of activity in Northeast Brazil, both the GSHAP model and the ABNT zoning have not taken into proper account some of the other highly active seismic zones, such as the Porto dos Gaúchos, Mato Grosso state, with recurrent activity since the occurrence of the largest recorded Brazilian event with magnitude 6.2 mb in 1955 \citep{barros_2009}, or the Central Amazon region where concentration of events in the last 50 years is consistent with the location of the possible historical earthquake of magnitude 7 \citep{veloso_2014}. For this reason, it is important to review and update the hazard estimates for Brazil, especially since the increase in the number of recordings of Brazilian earthquakes in the last three decades. In addition, the results presented here are part of the \acrfull{gem} supported SARA project (South America Risk Assessment).

Despite attempts to explain the origin of some of the seismic zones, such as stress concentrations from lithospheric thinning \citep{assumpcao_2004, azevedo_2015} or flexural effects \citep{zoback_1996, assumpcao_sacek_2013}, most of them have ill-defined boundaries. Attempts to produce seismic hazard maps \citep{dourado_2014} based on seismogenic zones are heavily dependent on subjective expert opinion, even more so in low seismic areas such as Brazil. As an important contribution to the new seismic hazard model for Brazil, here we briefly review recent methods of smoothing seismicity rates with application to Brazilian earthquakes.

\subsection{Previous studies in Brazil}

Despite some studies of seismicity rates and attempts at defining the main seismic zones \citep{berrocal_1984, berrocal_1996, assumpcao_et_al_2014}, Brazil has no seismic hazard map at national scale.

The Brazilian seismic hazard building code \citep{nbr_15421_2006} was based on the GSHAP \citep{giardini_1999} results for South America \citep{tanner_shepherd_1997, shedlock_tanner_1999}, which is the last published model for the whole region. However, some significant seismic areas (Porto dos Gaúchos \citep{barros_2009}), are not represented in the GSHAP-based building codes.

A significant number of earthquakes has been recorded in the last decades \citep{bsb_2014} allowing a better empirical and updated definition of the seismicity and hazard levels.

@@

\subsection{Smoothed seismicity}

Although the standard \citet{cornell_1968} and \citet{mcguire_1976} methodology often used to assess seismic hazard is using expert opinions to pre-defined seismic zones, the goal of this paper is to review a few smoothing seismicity methods based on \gls{kde} which results in a grid of smoothed seismicity without define any seismic zone polygon.

The \gls{kde} (or smoothed) seismicity grid allow to define a grid of seismic point sources. The seismic rate (\emph{a}-value) associated to each source is defined by the smoothed seismicity methods, but the other parameters, like \emph{b}-values, minimum and maximum magnitudes, rupture and depth distributions, must be assigned in other ways.

To perform the \gls{kde}, both its shape and bandwidth needs to be defined. Most kernels are Gaussian or power-laws or one of their variants. The main distinction between the methods described here is the bandwidth selection process.

\citet{frankel_1995} smoothing method applies a Gaussian \citet{nadaraya_1964}-\citet{watson_1964} smoothing technique with a \emph{fixed} smoothing-distance bandwidth to estimate the seismicity rate smoothing the 2D histogram counting. This version of the algorithm is already implemented on the \gls{hmtk} \citep{weatherill_2012, weatherill_2014_1}. There is also a simple variant of Frankel's method discussed by \citet{zechar_jordan_2010} which sum the contribution of each earthquake kernel on given cell. \citet{woo_1996} proposes a \emph{magnitude-dependent} kernel bandwidth based on the nearest neighbor mean distance in magnitude bins. \citet{helmstetter_2012} proposes a \emph{locally-adaptive} bandwidth, based on kernel estimations for the background seismicity rate in long-term forecasts on both space and time dimensions.

Using the background seismic rate computed to the forecast and assuming time-independence of that, the same assumption of PSHA, it is possible, preserving their differences \citep{marzocchi_2011}, to use these forecasted seismicity values on seismic sources characterization \citep{weatherill_pagani_2014}.

@@

To perform the hazard computation, the OpenQuake suite \citep{pagani_2014} was chosen. The modelling tool was the \gls{hmtk} also open available by \gls{gem} scientific board. The Frankel method was already implemented on the toolkit and the other two was implemented in the context of this paper and they are free available on a public repository.

@@

The specific purposes of this paper are: (i) review 3 distincts smoothing/zoneless methods to describe spatially the seismicity rate, (ii) update the seismic hazard studies at national scale by proposing a new hazard model,
(iii) use the GEM-OpenQuake tools to perform hazard calculation and (iv) add and share with the community two more new smoothing methods into the \gls{hmtk}.

To aim this, the Brazilian seismicity data is presented including some overview plots. Next the method's theory is presented follow by the specific decisions and optimizations performed on each method modelling. And last, the results, conclusions and further considerations are discussed.

@@

Today the catalog is distributed in two ways. One, which could be called \emph{raw}, contains all compiled information, such as: events outside the country which were felt in Brazil, errors in previous versions and also non-seismic events (known quarry blasts, sonic boom, etc). The other way could be called \emph{clean}, where all non-seismic and low quality events were removed. This last one was used in this paper and the total number of entries is [CALCULAR O NUMERO DE EVENTOS NO CATALOGO].

This catalog only cover crustal events. Andean subduction earthquakes deeper than 50km in the Brazil-Peru border was discarded. The epicenters of the catalog are shown in figure \ref{fig_seismicity}.

>>>>

\begin{abstract}
Intraplate seismicity is a hard problem to solve in tectonics. Brazil, as others intraplate environments, mainly during the lack of observed seismicity the seismic hazard assessment are also difficult and the methods applied are often close correlated to the geometries defined by the experts with valuable criteria.

Three smoothed seismicity techniques are reviewed and their application on construct models of Brazilian seismicity and its applications on long-term seismic hazard assessment in Brazil are also discussed. All these smoothing methods are based on a \gls{kde}. The main difference between them is the bandwidth selection process.

@@

All the hazard computations was performed on \gls{oq} \citep{pagani_2014} open source software from \acrfull{gem} Foundation. No focal mechanism information has been included or considered on these smoothed seismicity models. All hazard computations has been done using a single \gls{GMPE}. A more detailed study on \glspl{GMPE} selection should be developed further and it is beyond the scope of this work.

The results shown that all methods clearly most of the well known seismogenic regions and could be used in the Brazilian seismicity modeling. All methods improve the spatial resolution of the GSHAP hazard model, the only previously available for the region. These results should be considered on future reviews of Brazilian design building codes and procedures.

@@

\subsection{Intraplate seismicity and previous studies in Brazil}

Brazil occupies a large stable continental region with seismicity rate lower than Central and Eastern North America, Australia, India or Northern Europe \citep{giardini_1999,johnston_1996a}. Spanning most of the central region of the South American plate, Brazil has earthquakes above magnitude 6 mb with a return period of 50 years or so. Historical records are very short with almost no data before the XX century. Historical evidence of magnitude a 7 has been reported in the Amazon \citep{berrocal_1984,veloso_2014}. However, as in other intraplate settings, seismic hazard is not negligible due to the low seismic attenuation in the cold and thick lithosphere (magnitude 5 mb can be felt up to ~500 km away \citep{assumpcao_et_al_2014}) and special seismic hazard studies have been required for critical facilities such as nuclear power plants \citep{almeida_etal_2013}.

Very few studies related to seismic hazard, such as \citet{berrocal_1996} for SE Brazil, have been carried out so far and the first Brazilian seismic building code \citep{nbr_15421_2006} was based on the GSHAP \citep{giardini_1999} results for South America \citep{tanner_shepherd_1997, shedlock_tanner_1999}. Brazilian seismicity is concentrated in several seismic zones \citep{berrocal_1984, assumpcao_et_al_2014}. Except for a peak of activity in Northeast Brazil, both the GSHAP model and the ABNT zoning have not taken into proper account some of the other highly active seismic zones, such as the Porto dos Gaúchos, Mato Grosso state, with recurrent activity since the occurrence of the largest recorded Brazilian event with magnitude 6.2 mb in 1955 \citep{barros_2009}, or the Central Amazon region where concentration of events in the last 50 years is consistent with the location of the possible historical earthquake of magnitude 7 \citep{veloso_2014}. For this reason, it is important to review and update the hazard estimates for Brazil, especially since the increase in the number of recordings of Brazilian earthquakes in the last three decades. 
%In addition, the results presented here are part of the \acrfull{gem} supported SARA project (South America Risk Assessment).

Despite attempts to explain the origin of some of the seismic zones, such as stress concentrations from lithospheric thinning \citep{assumpcao_2004, azevedo_2015} or flexural effects \citep{zoback_1996, assumpcao_sacek_2013}, most of them have ill-defined boundaries. Attempts to produce seismic hazard maps \citep{dourado_2014} based on seismogenic zones are heavily dependent on subjective expert opinion, even more so in low seismic areas such as Brazil. As a contribution to the new seismic hazard model for Brazil, here we briefly review recent methods of smoothing seismicity rates with application to Brazilian earthquakes.

%Despite some studies of seismicity rates and attempts at defining the main seismic zones \citep{berrocal_1984, berrocal_1996, assumpcao_et_al_2014}, Brazil has no seismic hazard map at national scale.

%The Brazilian seismic hazard building code \citep{nbr_15421_2006} was based on the GSHAP \citep{giardini_1999} results for South America \citep{tanner_shepherd_1997, shedlock_tanner_1999}, which is the last published model for the whole region. However, some significant seismic areas (Porto dos Gaúchos \citep{barros_2009}), are not represented in the GSHAP-based building codes.

%A significant number of earthquakes has been recorded in the last decades \citep{bsb_2014} allowing a better empirical and updated definition of the seismicity and hazard levels.

@@

\subsection{Smoothed seismicity models}

Although the standard \citet{cornell_1968}-\citet{mcguire_1976} methodology often used to assess seismic hazard makes use of expert opinions to pre-defined seismogenic zones, one goal of this paper is to review a few smoothed seismicity methods based on \gls{kde} which results in a grid or a smoothed seismicity surface without define any seismic zone polygon (also known as \textit{zoneless} approaches).

The (smoothed) seismicity \gls{kde} techniques allow us to define another grid of seismic point sources. The seismicity rate (\emph{a}-value) associated to each source is defined by the smoothed seismicity methods.

To perform the \gls{kde}, both its shape and bandwidth needs to be defined. Most kernels are Gaussian, power-laws or one of their variants like Fisher kernels. The main distinction between the methods described here is the bandwidth selection process.

\citet{frankel_1995} smoothing method applies a Gaussian \citet{nadaraya_1964}-\citet{watson_1964} kernel smooth technique with a \emph{fixed smoothing-distance} bandwidth to estimate the seismicity rate by smoothing the 2D histogram from earthquakes counting. The actual version of the algorithm is already implemented on the \gls{hmtk} \citep{weatherill_2012, weatherill_2014_1} with a simple variant of Frankel's method discussed by \citet{zechar_jordan_2010} which sum the contribution of each earthquake kernel on given cell.

\citet{woo_1996} proposes a \emph{magnitude-dependent} kernel bandwidth based on the nearest neighbor mean distance in magnitude bins.

\citet{helmstetter_2012} proposes a \emph{locally-adaptive} bandwidth on both space and time dimensions for the background (time-independent) seismicity rate on their long-term forecasts experiments.

Using the long-term background seismic rate computed for forecast and assuming its time-independence, the same \gls{psha} assumption, it is possible, preserving their purposes differences \citep{marzocchi_2011}, to use these background seismicity rate values on seismic sources characterization \citep{weatherill_pagani_2014}.

\subsection{Information scores and \gls{ed}}

From these three smoothing methods, only \cite{helmstetter_2012} since its formalization use optimizations to define their model parameters. All the other methods are concept to use the whole catalog and some fixed parameter (e.g. Frankel's smoothing distance).

To compare the performance of these three models, the information scores and \gls{ed} \citep{molchan_1992, kagan_2007, kagan_2009} where computed as alternative to the tests suggested to the \gls{relm} forecasts experiments by \citet{schorlemmer_2007}.

On the evaluation phase all models learned from earthquakes into 1970-2004 observation time and were tested against the target realizations occurred between 2004 and 2014-01-01.

Scores from some Frankel-based models (from different smoothing distances) was computed and the best scores should give us an reasonable choice for this parameter.

HYBRID DECISION APPROACH BIRD/JORDAN[2013?]

@@

To perform the hazard computation, the \acrfull{oq} suite \citep{pagani_2014} was chosen. The modeling tool was the \gls{hmtk} also open available by \gls{gem} scientific board. The Frankel method was already implemented on the toolkit and two others was implemented in the context of this paper and they are free available on a public repository.

The other parameters, like \emph{b}-values, minimum and maximum magnitudes for the frequency-magnitude distributions, strike, dip and rate probabilities to the rupture plane, its top and bottom limits on the subsurface, magnitude-area scale relationships and also the depth distributions must be assigned in other ways to proceed the hazard computation and take advantages of the new developed \glspl{gmpe}.

In this work, to keep our focus on seismicity models as smoothed seismic sources, we perform hazard defining a single \gls{gmpe}  [EXPOSE WHICH ONE]  for a stable continental crust tectonic environment.

@@

The specific purposes of this paper are:

(i) review three distinct smoothing (zoneless) methods to spatially model the seismic rate,
(ii) update the seismic hazard studies at national scale by proposing a new simple hazard model,
(iii) use the \gls{gem}-\glsdesc{oq} tools to perform hazard calculation and 
(iv) add and share with the community two more new smoothing methods into the \gls{hmtk}.

To aim this, the Brazilian seismicity data is presented including some overview plots. Next the method's theory is given by the specific decisions and optimizations performed on each  method. And last, the results, conclusions and further considerations are discussed.

@@

Today the catalog is distributed in two ways. One, which could be called \emph{raw}, contains all compiled information, such as: events outside the country which were felt in Brazil, errors in previous versions and also non-seismic events (known quarry blasts, sonic boom, etc). The other way could be called \emph{clean}, where all non-seismic and low quality events were removed. This last one was used in this paper. There is 1941 earthquakes recorded from 1724 to the end of 2013.

This catalog only cover crustal events. Andean subduction earthquakes deeper than 50 km in the Brazil-Peru border was discarded as well earthquakes from the Mid-Atlantic ridge. The epicenters of the catalog are shown in figure \ref{fig_seismicity}.
